---
title: "Credit_Risk Model_Gradient_Descent"
output: html_document
---
```{r}
loan = read.csv("loan.csv" , na = "")
```

```{r}
library(tidyverse)
loan = loan %>%
        select(loan_status , loan_amnt , int_rate , grade , emp_length , home_ownership , 
               annual_inc , term)
loan
```

```{r}
sapply(loan , function(x) sum(is.na(x)))
```

```{r}
loan = loan %>%
        filter(!is.na(annual_inc) , 
               !(home_ownership %in% c('NONE' , 'ANY')) , 
               emp_length != 'n/a')
attach(loan)
```

Exploratory Data Analysis

```{r}
loan %>%
        count(loan_status) %>%
        ggplot(aes(x = reorder(loan_status , desc(n)) , y = n , fill = n)) + 
        geom_col() + 
        coord_flip() + 
        labs(x = 'Loan Status' , y = 'Count')
```

```{r}
loan = loan %>%
        mutate(loan_outcome = ifelse(loan_status %in% c('Charged Off' , 'Default') , 
                                     1, 
                                     ifelse(loan_status == 'Fully Paid' , 0 , 'No info')
                                     ))

barplot(table(loan$loan_outcome) , col = 'lightblue')
```

```{r}
loan2 = loan %>%
        select(-loan_status) %>%
        filter(loan_outcome %in% c(0 , 1))
```

```{r}
loan2$loan_outcome <- as.numeric(loan2$loan_outcome)
```

```{r}
plot(loan2$loan_amnt,loan2$annual_inc,col=as.factor(loan2$loan_outcome),xlab="Loan Amount",ylab="Annual Income")
```

```{r}
str(loan2)
```

```{r}
library(caret)
#Converting every categorical variable to numerical using dummy variables
dmy <- dummyVars("~ .", data = loan2,fullRank = T)
loan_transformed <- data.frame(predict(dmy, newdata = loan2))

#Checking the structure of transformed train file
str(loan_transformed)
```

```{r}
#Predictor variables
X <- as.matrix(loan_transformed[,c(-27)])
```

```{r}
#Add ones to X
X <- cbind(rep(1,nrow(X)),X)
```

```{r}
#Response variable
Y <- loan_transformed$loan_outcome
```


Our first step is to implement sigmoid function.
```{r}
#Sigmoid function
sigmoid <- function(z)
{
g <- 1/(1+exp(-z))
return(g)
}
```

Now we will implement cost function
```{r}
#Cost Function
cost <- function(theta)
{
m <- nrow(X)
g <- sigmoid(X%*%theta)
J <- (1/m)*sum((-Y*log(g)) - ((1-Y)*log(1-g)))
return(J)
}
```


Let’s test this cost function with initial theta parameters. We will set theta parameters equal to zero initially and check the cost.
```{r}
#Initial theta
initial_theta <- rep(0,ncol(X))

#Cost at inital theta
cost(initial_theta)
```


```{r}
str(X)
```


You will find cost is 0.693 with initial parameters. Now, our objective is to minimize this cost and derive the optimal value of the thetas.

```{r}
# Derive theta using gradient descent using optim function
theta_optim <- optim(par=initial_theta,fn=cost)

#set theta
theta <- theta_optim$par

#cost at optimal value of the theta
theta_optim$value
```

```{r}
prob <- sigmoid(t(c(1,30000,22.35,0,0,1,0,0,0,0,0,0,0,1,0,0,0,0,0,0,1,0,0,0,0,100000,0))%*%theta)
prob
```










